Rapid Convergence of The Born Series

### 2012-08-07 Thesis - Presentation/draft1.tex
Tuesday AUG 7th 2012 14:15

\documentclass[a4paper,10pt]{report}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{glossaries}
\usepackage[menubordercolor={0 0 0}]{hyperref}
\hypersetup{menubordercolor={0 0 0}}
%opening
\title{Rapid Convergence of The Born Series}
\author{Anupam Prasad Vedurmudi}
\date{July 2012}

@@

\section{Born Series - FEM Formulation}
We now use the definition of the Greens function in \eqref{GFdefinition} and the FEM formulation of the
Schrödinger equation, \eqref{schroedingerfem} to calculate the FEM representation of the Greens Function. 
We require that,
\begin{equation}\label{greenfemreq}
 G_{fem}(z)\psi=(z-E_0)^{-1}S\psi
\end{equation}
 Where, $\psi$ is an eigenstate of $H$ of energy $E_0$ and $z\neq E_0$ is a complex number. Thus, we define the Green's operator as,
\begin{equation}\label{greenfemdef}
 G_{fem}(z)=\left(z\mathbf{I}-HS^{-1}\right)^{-1}S
\end{equation}
 Where, $\mathbf{I}$ is the identity matrix. The free Greens function is defined similarly by setting
$H=B/2$. We will ultimately be testing the convergence of the matrix element, $\langle\mathbf{k}|G_{fem}(z)|\mathbf{k}\rangle$.
For later convenience, the potential matrix is also suitably modified as,
\begin{equation}\label{potmat}
 V_{fem}=S^{-1}VS^{-1}
\end{equation}

@@

Let the bound states (indexed by $a$) of the Hamiltonian $H$ be denoted by $|\psi_a\rangle$.
The potential is modified by the operator $\Gamma$ given by,
\begin{equation}\label{potmod}
  \Gamma=\lambda\displaystyle\sum_a |\psi_a\rangle\langle\psi_a|
\end{equation}
The rate of convergence of the Born series is controlled by the parameter $\lambda$. It is
seen that for weak potentials, $\lambda$ has very little influence on the convergence while
for stronger potentials which support several bound states, $\lambda$ seems to have
an optimal range for convergence.

The modified Hamiltonian and the modified free Hamiltonian are denoted by,
\begin{align}
 \widetilde{H}&=H+\Gamma\\
 \widetilde{H}_0&=H_0+\Gamma
\end{align}
The modified Greens functions $\widetilde{G}$ and $\widetilde{G}_0$ are defined similarly as
in \eqref{GFdefinition}. The modified Born Series for $\widetilde{G}$ is given by,

\begin{equation}\label{BornSeriesModCompact}
 \widetilde{G}=\displaystyle\sum_{n=0}^{\infty}\widetilde{G}_0\left(V\widetilde{G}_0\right)^n
\end{equation}
Physically, this means that the modified Hamiltonian has the same continuous spectrum but the discrete
spectrum has been shifted by $\lambda$.

\section{Example Problem - Scattering in 1D}
We now use the above methods in an example with scattering off a potential in 1D. Due to the simplicity of the problem,
it is possible to explicitly calculate the exact Green's operator by inverting the appropriate matrix according the formal
definition given in \eqref{GFdefinition}. We will examine the problem in two cases: 
\begin{itemize}
 \item In the first case, we have convergence with both the modified and unmodified Hamiltonian but at a faster rate with the modification.
 \item In the second case, the modified hamiltonian gives a convergent Born series whereas the unmodified one diverges. 
\end{itemize}

@@

The ``cosine eigenvectors'' are computed using \eqref{cosevec} and \eqref{freeschroedinger}.
The energies of the free wavefunctions are $E_n = n^2\pi^2/2L^2$ for $n=0,\ldots,N$ and $L=10$.
The Green's operator, $G(z)$ is calculated for $z=E_n+i\epsilon$,
where $\epsilon$ is a small positive number. The purpose of $\epsilon$ is to avoid 
the singularity of the Green's operator on the positive real axis. We first encountered it in \eqref{GFdefinition} where
we originally took $\epsilon\rightarrow 0$. In the numerical example, however, epsilon is fixed to be a small
number which is larger than the energy spacing between the bound states close to the origin.

\subsection{First Case}
%\begin{center}
\begin{figure}
\includegraphics[width=320pt, height=145pt]{series1a.png}
\includegraphics[width=320pt, height=145pt]{series1b.png}
\includegraphics[width=320pt, height=145pt]{series1c.png}
\caption[\textwidth]{First Case - Weak Finite Well}
\end{figure}
%\end{center}

In the first case, we define the potential as a finite well of strength $0.01$ with support in $4.5 \leq x \leq 5.5$,
\begin{equation}\label{potentialdef1}
V = \begin{cases}
-0.01 &\mbox{} 4.5\leq x\leq 5.5\\
0 &\mbox{ otherwise}
\end{cases} 
\end{equation}
The Born Series is calculated for $15$ iterations. The Hamiltonian is modified using the operator defined in \eqref{potmod} with
$\lambda=1$.  We also set $\epsilon=10^{-3}$.

We now plot the elements of the Born Series against the number of iterations. In order
to compare the modified and unmodified cases, we plot them side by side with the modified case on the left and the unmodified one
on the right. We follow this convention in the other next case as well.

\subsection{Second Case}
The potential is give by,
\begin{equation}\label{potentialdef3}
V = -100\mathtt{exp}(-(x-3.5)^2/0.5)-100\mathtt{exp}(-(x-6.5)^2/0.5)
\end{equation}
We set $\epsilon=10^{-6}$ and $\lambda=100$.
\begin{figure}[hb]
\centering
\includegraphics[width=225pt, height=150pt]{potential1d2.png}
\caption[\textwidth]{Test Potential 2}
\end{figure}
\begin{figure}
\includegraphics[width=320pt, height=145pt]{series3a.png}
\end{figure}
\begin{figure}[ht]
%\includegraphics[width=320pt, height=145pt]{series3a.png}
\includegraphics[width=320pt, height=145pt]{series3b.png}
\includegraphics[width=320pt, height=145pt]{series3c.png}
\caption[\textwidth]{Gaussian Like Potential}
\end{figure}

We see that for some energies the Born Series in the modified case converges to the exact value but diverges in the original case. We also note that
for certain energies, the Born Series diverges in both cases but the divergence in the unmodified case is slower.

\chapter{Conclusion}

\appendix
\addcontentsline{toc}{chapter}{Appendices}
%\chapter{Calculation of The Scattering Cross-section}\label{crosssection}
\chapter{Python Implementation}\label{pythonimp}
The main object of interest is the \verb=Axis= class, which is defined in the file \verb=axis.py=.

>>>>

\documentclass[a4paper,10pt]{report}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{glossaries}
\usepackage[menubordercolor={0 0 0}]{hyperref}
\hypersetup{menubordercolor={0 0 0}}
\usepackage{float}
%opening
\title{Rapid Convergence of The Born Series}
\author{Anupam Prasad Vedurmudi}
\date{July 2012}

@@

\section{Born Series - FEM Formulation}
We now use the definition of the Greens function in \eqref{GFdefinition} and the FEM formulation of the
Schrödinger equation, \eqref{schroedingerfem} to calculate the FEM representation of the Greens Function. 
We require that,
\begin{equation}\label{greenfemreq}
 G_{fem}(z)\psi=(z-E_0)^{-1}S\psi
\end{equation}
 Where, $\psi$ is an eigenstate of $H$ of energy $E_0$ and $z\neq E_0$ is a complex number. Thus, we define the Green's operator as the matrix,
\begin{equation}\label{greenfemdef}
 G_{fem}(z)=\left(z\mathbf{I}-HS^{-1}\right)^{-1}S
\end{equation}
 Where, $\mathbf{I}$ is the identity matrix. The free Greens function is defined similarly by setting
$H=B/2$. We will ultimately be testing the convergence of the matrix element, $\langle\mathbf{k}|G_{fem}(z)|\mathbf{k}\rangle$.
For later convenience, the potential matrix is also suitably modified as,
\begin{equation}\label{potmat}
 V_{fem}=S^{-1}VS^{-1}
\end{equation}

@@

Let the bound states (indexed by $a$) of the Hamiltonian $H$ be denoted by $|\varphi_a\rangle$.
The potential is modified by the operator $\Gamma$ given by,
\begin{equation}\label{potmod}
  \Gamma=\displaystyle\sum_a |\varphi_a\rangle\langle\varphi_a|
\end{equation}
The above operator is multiplied by a  positive factor $\lambda$ which controls
the rate of convergence of the Born series. It is
seen that for weak potentials, $\lambda$ has very little influence on the convergence while
for stronger potentials which support several bound states, $\lambda$ seems to have
an optimal range for convergence.

The modified Hamiltonian and the modified free Hamiltonian are denoted by,
\begin{align}
 \widetilde{H}&=H+\lambda\Gamma\\
 \widetilde{H}_0&=H_0+\lambda\Gamma
\end{align}
The modified Greens functions $\widetilde{G}$ and $\widetilde{G}_0$ are defined similarly as
in \eqref{GFdefinition}. The modified Born Series for $\widetilde{G}$ is given by,

\begin{equation}\label{BornSeriesModCompact}
 \widetilde{G}=\displaystyle\sum_{n=0}^{\infty}\widetilde{G}_0\left(V\widetilde{G}_0\right)^n
\end{equation}
Physically, this means that the modified Hamiltonian has the same continuous spectrum but the discrete
spectrum has been shifted by $\lambda$.
\subsubsection{Calculating Original Green's Operator}
The original Green's operator is related to the modified Green's operator through the following
Lippmann-Schwinger equation,
\begin{equation}\label{lippmannschwinger2}
 G=\widetilde{G}-\lambda\widetilde{G}\Gamma G
\end{equation}
We can then use the fact that,
\begin{equation}
 \widetilde{\Gamma}\equiv G(z)\Gamma=\Gamma G(z) = \displaystyle\sum_a \frac{|\varphi_a\rangle\langle\varphi_a|}{z-E_a} 
\end{equation}
where, $E_a$ is the energy of the bound state $|\varphi_a\rangle$. The above operator does not diverge
since $Re(z)>0$ and $E_a<0$. Denoting the $n^{th}$ order
approximation of $\widetilde{G}$ in \eqref{BornSeriesModCompact} as $\widetilde{G}^{(n)}$ we get,
\begin{equation}\label{Gapprox}
 G^{(n)}\approx \widetilde{G}^{(n)}+\lambda\widetilde{G}^{(n)}\widetilde{\Gamma}
\end{equation}

\subsubsection{$\lambda\rightarrow\infty$ Limit}
Before we go on to examples where we apply the orthogonal projection method, we will briefly review the behaviour of
the above technique in the $\lambda\rightarrow\infty$ limit as described in \cite{KukPom78}. The Lippmann-Schwinger equation \eqref{lippmannschwinger2} can also be written as,
\begin{equation}\label{lippmannschwinger3}
 \widetilde{G}=G+\lambda G\Gamma\widetilde{G}
\end{equation}
Where we have obtained an iterative expression for $\widetilde{G}$ in terms of $G$. From this we immediately obtain,
\begin{equation}
 \widetilde{G}=G+G\Gamma\lambda\left(\Gamma\left(I-\lambda G\right)\Gamma\right)^{-1}\Gamma G
\end{equation}
The admixture nonorthogonal to $\Gamma$ is,
\begin{equation}
 \Gamma\widetilde{G}=\Gamma\left(\Gamma\left(I-\lambda G\right)\Gamma\right)^{-1}\Gamma G \displaystyle\xrightarrow[\lambda\rightarrow\infty]{} 0,
\end{equation}
i.e., at $\lambda\rightarrow\infty$, the $\widetilde{G}$ is orthogonal to the subspace of the bound states
\begin{equation}
 \widetilde{G}=G-G\Gamma\left(\Gamma G\Gamma\right)^{-1}\Gamma G
\end{equation}
Where the inverse is taken only in the subspace of the bound states. The above expression thus reduces to,
\begin{equation}
 \widetilde{G}\rightarrow \left(I-\Gamma\right)G = G-\displaystyle\sum_i\frac{|\varphi_i\rangle\langle\varphi_i|}{E-E_i}
\end{equation}
Effectively, this means that the poles corresponding to the bound states have been eliminated from the original
Green's operator. This limit was not applied in the numerical examples because of the high computational 
overhead required. The above expression would have to be calculated separately for every energy, whereas for a finite 
$\lambda$, only one call is required for all scattering energies.
\section{Example Problem - Scattering in 1D}
We now use the above methods in an example with scattering off a potential in 1D. Due to the simplicity of the problem,
it is possible to explicitly calculate the exact Green's operator by inverting the appropriate matrix according the formal
definition given in \eqref{GFdefinition}. We will examine the problem in two cases: 
\begin{itemize}
 \item In the first case, we have convergence with both the modified and unmodified Hamiltonian but at a faster rate with the modification.
 \item In the second case, the modified hamiltonian gives a convergent Born series whereas the unmodified one diverges. 
\end{itemize}

@@

The ``cosine eigenvectors'' are computed using \eqref{cosevec} and \eqref{freeschroedinger}.
The energies of the free wavefunctions are $E_n = n^2\pi^2/2L^2$ for $n=0,\ldots,N$ and $L=10$.
The Green's operator, $G(z)$ is calculated for $z=E_n+i\epsilon$,
where $\epsilon$ is a small positive number. The purpose of $\epsilon$ is to avoid 
the singularity of the Green's operator on the positive real axis. We first encountered it in \eqref{GFdefinition} where
we originally took $\epsilon\rightarrow 0$. In the numerical example, however, epsilon is fixed to be a small
number which is larger than the energy spacing between the bound states close to the origin.
\subsection{First Case}
In the first case, we define the potential as a finite well of strength $0.01$ with support in $4.5 \leq x \leq 5.5$,
\begin{equation}\label{potentialdef1}
V = \begin{cases}
-0.01 &\mbox{} 4.5\leq x\leq 5.5\\
0 &\mbox{ otherwise}
\end{cases} 
\end{equation}
The Born Series is calculated for $15$ iterations. The Hamiltonian is modified using the operator defined in \eqref{potmod} with
$\lambda=1$.  We also set $\epsilon=10^{-3}$.

\subsubsection{Description of Plot}
We now plot the elements of the Born Series against the number of iterations for different scattering energies, $E_n$. In order
to compare the modified and unmodified cases, we plot them side by side with the modified case on the left and the unmodified one
on the right. The x-axis denotes the number of iterations of the Born series, and the y-axis the matrix element of the
Green's operator, $\langle\mathbf{k}|G(E_n+i\epsilon)|\mathbf{k}\rangle$ ($|\mathbf{k}\rangle$ defined in \eqref{cosevec}) in log-scale. 
We follow this convention in the next case as well.

\begin{figure}[hb]
\includegraphics[width=320pt, height=135pt]{series1a.png}
\includegraphics[width=320pt, height=135pt]{series1d.png}
\end{figure}
\begin{figure}[ht]
%\includegraphics[width=320pt, height=135pt]{series1a.png}
%\includegraphics[width=320pt, height=135pt]{series1d.png}
\includegraphics[width=320pt, height=140pt]{series1b.png}
\includegraphics[width=320pt, height=140pt]{series1c.png}
\caption[\textwidth]{First Case - Weak Finite Well}
\end{figure}

\subsection{Second Case}
The second test potential is a mixture of a finite-well and a gaussian. Its plot is 
given below. We set $\epsilon=10^{-6}$ and $\lambda=100$.
\begin{figure}[H]
\centering
\includegraphics[width=215pt, height=135pt]{potential1d2.png}
\caption[\textwidth]{Test Potential 2}
\end{figure}
%\begin{figure}
%\includegraphics[width=320pt, height=135pt]{series3a.png}
%\end{figure}
\begin{figure}[ht]
\includegraphics[width=320pt, height=135pt]{series3a.png}
\includegraphics[width=320pt, height=135pt]{series3b.png}
\includegraphics[width=320pt, height=135pt]{series3c.png}
\includegraphics[width=320pt, height=135pt]{series3d.png}
\caption[\textwidth]{Finite Well + Gaussian}
\end{figure}

\chapter{Conclusion}
We see that for some energies the Born Series in the modified case converges to the exact value but diverges in the original case. We also note that
for certain energies, the Born Series diverges in both cases but the divergence in the unmodified case is slower.
%\appendix
%\addcontentsline{toc}{chapter}{Appendices}
%\chapter{Calculation of The Scattering Cross-section}\label{crosssection}
%\chapter{Python Implementation}\label{pythonimp}
%The main object of interest is the \verb=Axis= class, which is defined in the file \verb=axis.py=.
